Ace: An Actively-Typed Language and Compilation Environment for High-Performance Computing

### 2013-04-23 ace/ace.tex
Updated Ace intro

Computer-aided simulations and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}. Most of this software targets desktop hardware, while about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. To fully harness the power of these platforms, however, these {\em professional end-user developers} \cite{segal2007some} must increasingly write parallel programs.

Professional end-users today generally use dynamically-typed scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where these interpretted languages are too slow, they will typically call into code written in a statically-typed, low-level language like C or Fortran that uses low-level parallel abstractions like pthreads or MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use properly, even for expert programmers.

The research community has responded with dozens of language features and parallel programming abstractions that aim to  more equitably balance various concerns relating to performance, portability, verifiability and ease-of-use. Unfortunately, professional end-users have become skeptical that new approaches that originate in the research community are practical. This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance high-performance computing.

We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language. Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
\begin{itemize}
\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
\end{itemize}

The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style. Empirical evidence also supports this notion that developers prefer first-class support for a parallel abstraction over a library-based solution, if other factors are held constant \cite{ppfirstclass}.

Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}. These are significant enough that programming abstractions that are distributed as libraries for a well-established language, such as MapReduce, Global Arrays and MPI, have prevailed in practice, despite the limited expressivity of library-based abstractions. This is balanced by several factors: developers can progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize arbitrary combinations of  abstractions within a single program without needing to navigate the friction inherent to language boundaries.

Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}. Based on the notion of active libraries \cite{activelibraries}, \ATT~is designed to allow developers to  extend the semantics, but not the syntax, of a language. The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe. We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}. Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.

Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today. It is based on the Python programming language, which enjoys broad adoption in this community. Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.

>>>>

Computer-aided simulation and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}. Most of this software targets desktop hardware, while about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. To fully harness the power of these platforms, however, these so-called {\em professional end-user developers} \cite{segal2007some} must increasingly write parallel programs.

Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language like C or Fortran that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use properly because verifying code written with them remains a difficult (and often ill-specified) problem.

Researchers and domain experts often respond to these challenges by proposing novel language features that aim to strike an intermediate balance between \textbf{performance}, \textbf{verifiability\-}, \textbf{portability} and \textbf{ease-of-use}.
Unfortunately, professional end-users rarely adopt new languages, even when they have been actively developed. Indeed, many end-users have become skeptical that novel approaches that originate in the research community can be  practical. This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although it may seem paradoxical, the ubiquity of this sentiment demands direct examination by researchers proposing novel abstractions and languages for professional end-users in HPC.

We suggest three mutually-related {design criteria} that, unlike those in bold above, many languages and language-integrated abstractions have failed to adequately consider: \textbf{continuity}, \textbf{extensibility} and \textbf{interoperability}. These criteria encompass the intuitions that new abstractions will not be adopted in a vacuum, that programming systems must  support change, and that interacting components of an application or workflow should be able to make use of different abstractions naturally and without the possibility of conflict arising at their interface boundaries.
%that this is far from being a superfluous issue. These criteria can influence the core design of a language or abstraction.

Ace is a programming language intended to be used by professional end-users as well as researchers in areas related to high-performance computing and computational science. The primary contribution of this paper consists of describing several novel language mechanisms and novel combinations of existing mechanisms that Ace uses to cohesively address the design criteria above. We anticipate that many of these mechanisms and choices will also influence other languages in the future. 
Following a principled design methodology, we justify each design decision by appeal to this full set of design criteria and citations to relevant literature. These criteria and methodology constitute another contribution of this work that we hope  will lead to fewer \emph{ad hoc} designs and serve as a foundation for evaluating competing designs in HPC and related domains in the future.

We begin in Section \ref{highlevel} with a high-level overview of the Ace language design and the reasoning behind the choices that were made. In Section \ref{usage} we provide further details by way of concrete examples employing Ace for low-level GPU programming, based on an internalization of the OpenCL programming language as an Ace library. Ace can be used both as a standalone language via the \verb|acec| compiler and as an interactive compilation environment. This latter mode of use enables just-in-time specialization and direct invocation of Ace functions from within Python scripts, as well as deep integration between the \verb|numpy| library, the OpenCL host API and the Ace compiler.

In Section \ref{att}, we describe the extensible type system, which is based on the novel idea of reifying types as metalanguage objects. This mechanism, called \emph{active typing}, was used to implement the OpenCL library. 
The operations associated with \emph{active types} are given meaning by translation to a user-defined backend. We show in Section \ref{translation} how this allows us to give semantics to OpenCL constructs by translation to OpenCL itself, as well as to CUDA and C99, increasing the operational flexibility of the OpenCL standard. Then in Section \ref{usecases}, we  continue with several examples of higher-level parallel abstractions that can be implemented in Ace. In other systems, these have required deriving a new variant of an existing language or a new language, both strategies that violate at least two of the three new criteria above. In Ace, these exist within libraries and can safely be used within a single program.

To demonstrate that Ace can be used for realistic scientific workloads, we briefly describe in Section \ref{casestudy} a case study that used Ace as the foundation for a neurobiological simulation framework capable of orchestrating the execution of thousands of realizations of a large stochastic neural circuit model on clusters of GPUs. 
We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of some limitations of Ace at the time of writing, as well as a discussion of future work building on both Ace itself as well as the concept of active typing central to Ace.

\section{The Design of Ace}\label{highlevel}
It is now widely-acknowledged that static type systems are fundamental tools for performance and verifiability, both critical criteria in HPC, so Ace is fundamentally a statically-typed language. 
To achieve continuity, Ace is implemented as a library in the ubiquitous scripting language Python. The Ace grammar is precisely that of Python, and it share's Python's packaging system, so any tools that work with Python files and packages can immediately be used with Ace.

Another important tool that has become increasingly prominent is the use of code generation techniques to optimize code and orchestrate scientific workflows, so Ace includes a powerful

\begin{itemize}
\item Types are fundamental tools for both performance and verifiability.
\item Programmers in high-performance computing tend to be wary of abstraction mechanisms that necessitate run-time overhead.
\item Compile-time code generation is often used to optimize code and orchestrate scientific workflows.
\item Scripting languages, such as Python, are particularly appreciated for their ease-of-use.
\item Library-based abstractions are more highly-adopted, but language-integrated abstractions are easier to use \cite{xxx}.
\end{itemize}

\section{Usage of Ace}\label{usage}

\section{Active Typechecking and Translation}\label{att}
\subsection{Active Translation}\label{translation}
\subsection{Use Cases}\label{usecases}
\section{Case Study}\label{casestudy}
\section{Related Work}\label{related}
\section{Discussion}\label{discussion}
%We argue that many of the barriers to adoption faced by new abstractions stem from the fact that new abstractions are too-often developed and distributed together with a new programming language. Some prominent cases where such a {\em language-oriented approach} \cite{languageoriented} has been followed include:
%\begin{itemize}
%\item UPC, CAF, X10 and Chapel, built around first-class support for a partitioned global address space (PGAS) programming model; 
%\item Go, Erlang, Charm++, Occam and Axum, built around first-class support for message passing; and
%\item CUDA, OpenCL, Cilk and NESL, built around first-class support for data parallelism.
%\end{itemize}
%
%The benefits of taking this approach are clear: first-class language, compiler and run-time support for a programming abstraction can improve performance and enable a more natural programming style. Empirical evidence also supports this notion that developers prefer first-class support for a parallel abstraction over a library-based solution, if other factors are held constant \cite{ppfirstclass}. 
%
%Unfortunately, adopting a specialized research language comes with a host of ancillary challenges  \cite{adoptionTR,socioplt}. These are significant enough that programming abstractions that are distributed as libraries for a well-established language, such as MapReduce, Global Arrays and MPI, have prevailed in practice, despite the limited expressivity of library-based abstractions. This is balanced by several factors: developers can progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize arbitrary combinations of  abstractions within a single program without needing to navigate the friction inherent to language boundaries. 
%
%Ace aims to bridge the language-oriented and library-oriented approaches by allowing developers to define and distribute a broad range of new {\em first-class} types and their operations directly within user libraries, using a mechanism called {\em active typechecking and translation (\ATT)}. Based on the notion of active libraries \cite{activelibraries}, \ATT~is designed to allow developers to  extend the semantics, but not the syntax, of a language. The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe. We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}. Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
%
%Although the \ATT~mechanism, a key feature of Ace, is designed primarily for use by researchers and domain experts, Ace is intended to be highly practical and ready for adoption by professional end-user developers today. It is based on the Python programming language, which enjoys broad adoption in this community. Indeed, the language itself is implemented entirely as a Python library, and many existing tools, such as editor modes, documentation tools and style checkers, can be used without modification.
%Ace functions are generated programmatically and can be called directly from Python, with numeric arrays as arguments, so developers can integrate Ace immediately within their existing workflows for portions of code that may benefit from static compilation and support for parallelism.
%We demonstrate the practicality of Ace in Section \ref{clegans} by describing an Ace-based framework that supports specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs.
